---
title: "Assignment_3_Stats_II_Group_22"
author:
- "Maik David Klaiber (22-482-426)"
- "Jonas Weber (24-621-484)"
- "Yann Rochaix (24-622-276)"
date: "`r format(Sys.Date(), '%d.%m.%Y')`"
output: html_document
---
```{r setup, include =FALSE}
library(tidyverse)
library(summarytools)
library(skimr)
```

## Exercise 1

Since $Q_i = X_i - 10$, substituting into 
$Y_{i,1} = \delta_0 + \delta_1 Q_i + u_i$ gives:

$$
Y_{i,1} = \delta_0 + \delta_1 (X_i - 10) + u_i
$$

Expanding the brackets and rearranging:

$$
Y_{i,1} = (\delta_0 - 10\delta_1) + \delta_1 X_i + u_i
$$

Comparing with $Y_{i,1} = \beta_0 + \beta_1 X_i + u_i$ implies:

$$
\delta_1 = \beta_1
$$

Therefore, the OLS estimators satisfy:

$$
\hat{\delta}_1 = \hat{\beta}_1
$$

#### What happens to the OLS coefficients, when a constant is added or subtracted?

The slope coefficient does not change:
$$
\hat{\delta}_1 = \hat{\beta}_1
$$

To see how the intercept changes, substitute $Q_i = X_i - c$ into the model:
$$
Y_i = \hat{\delta}_0 + \hat{\delta}_1 (X_i - c) + u_i
$$

Rearranging:
$$
Y_i = (\hat{\delta}_0 - c\hat{\delta}_1) + \hat{\delta}_1 X_i + u_i
$$

Comparing with $Y_i = \hat{\beta}_0 + \hat{\beta}_1 X_i + u_i$ yields:
$$
\hat{\beta}_0 = \hat{\delta}_0 - c\hat{\delta}_1
$$

Solving for the intercept:
$$
\hat{\delta}_0 = \hat{\beta}_0 + c\hat{\beta}_1
$$

Thus, subtracting a constant shifts the regression line but does not affect its slope.

The relationship between $\hat{\delta}_0$ and $\hat{\beta}_0$ remains unchanged and is given by the deterministic linear relationship:
$$
\hat{\delta}_0 = \hat{\beta}_0 + c\hat{\beta}_1
$$

## Exercise 2

#### (i) Importing, exploring and cleaning the dataset

```{r, results='hide'}
df <- read.csv("distances_birthtown_up.csv")
skim_without_charts(df)

``` 
In `p0` (minimum value), we see values such as -999999 in`income`, `income_parent1` and `income_parent2` which need to be replaced by `NA`. Also in the variable `travel_time` there are some negative values, which can't be real we also replace them with `NA`.
```{r, results='hide'}
df$income[df$income == -9999999] <- NA
df$income_parent1[df$income_parent1 == -9999999] <- NA
df$income_parent2[df$income_parent2 == -9999999] <- NA

df$travel_time[df$travel_time<0] <- NA
#We create the income_parents variable
df$income_parents <- df$income_parent1 + df$income_parent2
```

<br>

#### (ii) Writing the Regression Equation

$$medicine_i= \beta_0 + \beta_1travel\_time_i + \beta_2 income\_parents_i + u_i$$
```{r, message=F}
# Converting medicine to binary from text string
df$medicine_bin <- ifelse(df$medicine == "Yes", 1, 0)

model1 <- lm(medicine_bin ~ travel_time + income_parents, data = df)
summary(model1)
```
**Interpretation of the coefficient parameters**



- $\hat{\beta}_0$ (Intercept): The predicted probability of working in medicine for an individual with zero travel time and zero parental income.
- $\hat{\beta}_1$ (Travel time): A one-hour increase in travel time to Country B from town of birth is associated with a $\hat{\beta}_1$ change in the probability of working in medicine, holding parental income constant.
- $\hat{\beta}_2$ (Income parents): A one-unit ($1,000) increase in combined parental income is associated with a $\hat{\beta}_2$ change in the probability of working in medicine, holding travel time constant.
<br><br>
We assume that MLR.1-MLR.5 hold.

<br>

#### (iii) Comparison of the two regressors
From the previous sub exercise, this model accounts for both travel time and parental income:

$$
medicine_i = \beta_0 + \beta_1 \, travel\_time_i + \beta_2 \, income\_parents_i + u_i
$$

From the current sub exercise, this model focuses solely on the relationship with parental income:

$$
medicine_i = \delta_0 + \delta_1 \, income\_parents_i + v_i
$$
```{r}
model2 <- lm(medicine_bin ~ income_parents, data = df)

summary(model1)
summary(model2)
```
The coefficient on `income_parents` differs substantially across models: 
$\hat{\beta}_2 = 0.0002$ ($p = 0.141$) in model (ii) versus 
$\hat{\delta}_1 = 0.0008$ ($p < 0.001$) in model (iii).

At the 5\% significance level ($\alpha = 0.05$), the coefficient is not statistically significant in model (ii), but is statistically significant in model (iii). 
The magnitude of the coefficient is also economically different, as it is approximately four times larger in the simple regression.

This pattern is consistent with omitted variable bias. In particular, `travel_time` has a negative effect on the probability of working in medicine. 
Omitted variable bias arises if the omitted variable is correlated with the included regressor, i.e.
$$
\text{Cov}(travel\_time, income\_parents) \neq 0
$$

A plausible assumption is that individuals with higher parental income tend to live closer (i.e., have lower `travel_time`), implying a negative correlation between `travel_time` and `income_parents`. 
Under this assumption, omitting `travel_time` biases the coefficient on `income_parents` upward.

Thus, the difference in coefficients across models can be explained by the omission of a relevant variable.
<br>

#### (iv) Is `region` a better variable than `travel_time`?
Using `region` instead of `travel_time` would likely answer our research question less well. travel_time is a continuous variable that directly measures proximity to Country B's high-income medical industry, capturing meaningful variation across individuals.<br>
`region`, by contrast, is a categorical variable with codes 1–15 that groups individuals into broad geographic areas. This aggregation discards within-region variation in actual travel times — two individuals in the same region may face very different travel times to Country B. Furthermore, `travel_time` has a direct economic interpretation (each additional hour reduces the probability of working in medicine by 4.9 percentage points), whereas region codes have no inherent ordering or magnitude. Therefore, `travel_time` is a more precise and interpretable proxy for geographic proximity to the medical labour market.<br>

<br>

#### (v) Test whether `income_parents` is correlated with `travel_time`

To test whether `travel_time` and `income_parents` are correlated, 
we estimate the following regression:

$$
income\_parents_i = \gamma_0 + \gamma_1 travel\_time_i + e_i
$$

The null and alternative hypotheses are:

- $H_0: \gamma_1 = 0$ (no correlation between travel time and parental income)
- $H_1: \gamma_1 \neq 0$ (correlation exists)

The t-statistic is constructed as:

$$
t = \frac{\hat{\gamma}_1 - 0}{SE(\hat{\gamma}_1)}
$$

We test at the 5\% significance level ($\alpha = 0.05$).


```{r}
model_corr <- lm(income_parents ~ travel_time, data = df)
summary(model_corr)
```

**Conclusion** 

The estimated coefficient is $\hat{\gamma}_1 = -17.0877$, 
with a t-statistic of $-58.52$ and a p-value $< 0.001$. 
At the 5\% significance level ($\alpha = 0.05$), we reject $H_0$ and conclude that 
`travel_time` and `income_parents` are statistically significantly negatively correlated. Alternatively, comparing the t-statistic to the critical value, we have 
$|t| = 58.52 > t_{crit} \approx 1.96$ (for $\alpha = 0.05$), 
leading to the same conclusion of rejecting $H_0$. 

This provides evidence that `travel_time` is correlated with `income_parents`, 
supporting the omitted variable bias argument in part 2(iii).

<br>

#### (vi) Test whether `travel_time` and `income_parents` are jointly significant

We consider the regression model:

$$
medicine_i = \beta_0 + \beta_1 \, travel\_time_i + \beta_2 \, income\_parents_i + u_i
$$

We test whether `travel_time` and `income_parents` are jointly significant.

The null and alternative hypotheses are:

- $H_0$: $\beta_1 = 0$ and $\beta_2 = 0$  
  (no joint effect)

- $H_1$: $\beta_1 \neq 0 \ \text{or} \ \beta_2 \neq 0$  
  (At least one of the coefficients is different from zero i.e. joint significance)

We use an F-test comparing the unrestricted model (including both variables) to the restricted model (excluding them). The test statistic is:

$$
F = \frac{(SSR_r - SSR_{ur})/q}{SSR_{ur}/(n - k - 1)}
$$

where $q = 2$ is the number of restrictions. The test measures how much the model fit worsens when the restrictions are imposed.

The decision is based on comparing the F-statistic to the critical value at the chosen significance level. If $F > F_{crit}$, we reject $H_0$, implying that at least one coefficient is different from zero and the variables are jointly significant.

