---
title: "Assignment_3_Stats_II_Group_22"
author:
- "Maik David Klaiber (22-482-426)"
- "Jonas Weber (24-621-484)"
- "Yann Rochaix (24-622-276)"
date: "`r format(Sys.Date(), '%d.%m.%Y')`"
output: html_document
---
```{r setup, include =FALSE}
library(tidyverse)
library(summarytools)
library(skimr)
```

## Exercise 1

Since $Q_i = X_i - 10$, substituting into $Y_{i,1} = \delta_0 + \delta_1 Q_i + u_i$ gives:
$$Y_{i,1} = \delta_0 + \delta_1(X_i - 10) + u_i = \delta_0 - 10\delta_1 + \delta_1 X_i + u_i$$

Comparing and matching coefficients on $X_i$ with the original model: $Y_{i,1} = \beta_0 + \beta_1 X_i + u_i$, we get:
$$\hat{\delta}_1 = \hat{\beta}_1$$
And when matching the constants: $\hat{\delta}_0 = \hat{\beta}_0 + 10\hat{\beta}_1$, then rearranging:
$$\hat{\delta}_0 = \hat{\beta}_0 + 10\hat{\beta}_1$$

#### What happens to the OLS coefficients, when a constant is added or subtracted?
The slope does not change $\hat{\delta}_1 = \hat{\beta}_1$ when subtracting a constant from a regressor. 

Only the intercept will change $\hat{\delta}_0 = \hat{\beta}_0 + 10\hat{\beta}_1$ when the regressor gets transformed. 

## Exercise 2

#### Importing, exploring and cleaning the dataset

```{r, results='hide'}
df <- read.csv("distances_birthtown_up.csv")
skim_without_charts(df)

``` 
In `p0` (minimum value), we see values such as -999999 in`income`, `income_parent1` and `income_parent2` which need to be replaced by `NA`. Also in the variable `travel_time` there are some negative values, which can't be real we also replace them with `NA`.
```{r, results='hide'}
df$income[df$income == -9999999] <- NA
df$income_parent1[df$income_parent1 == -9999999] <- NA
df$income_parent2[df$income_parent2 == -9999999] <- NA

df$travel_time[df$travel_time<0] <- NA
#We create the income_parents variable
df$income_parents <- df$income_parent1 + df$income_parent2
```

#### Writing the Regression Equation

$$medicine_i= \beta_0 + \beta_1travel\_time_i + \beta_2 income\_parents_i + u_i$$
```{r, message=F}
# Converting medicine to binary from text string
df$medicine_bin <- ifelse(df$medicine == "Yes", 1, 0)

model1 <- lm(medicine_bin ~ travel_time + income_parents, data = df)
summary(model1)
```
#### Interpretation of the coefficient parameters
- $\hat{\beta}_0$ (Intercept): The predicted probability of working in medicine for an individual with zero travel time and zero parental income.
- $\hat{\beta}_1$ (Travel time): A one-hour increase in travel time to Country B from town of birth is associated with a $\hat{\beta}_1$ change in the probability of working in medicine, holding parental income constant.
- $\hat{\beta}_2$ (Income parents): A one-unit ($1,000) increase in combined parental income is associated with a $\hat{\beta}_2$ change in the probability of working in medicine, holding travel time constant.









