---
title: "Assignment_3_Stats_II_Group_22"
author:
- "Maik David Klaiber (22-482-426)"
- "Jonas Weber (24-621-484)"
- "Yann Rochaix (24-622-276)"
date: "`r format(Sys.Date(), '%d.%m.%Y')`"
output: html_document
---
```{r setup, include =FALSE}
library(tidyverse)
library(summarytools)
library(skimr)
```

## Exercise 1

We observe the variables $Y_{i,1}$, $Y_{i,2}$, and $X_i$, and estimate the following simple linear regressions:

$$
Y_{i,1} = \beta_0 + \beta_1 X_i + u_i
$$

$$
Y_{i,2} = \alpha_0 + \alpha_1 X_i + v_i
$$

Next, we define the transformed variable:

$$
Q_i = X_i - 10
$$
Substituting $Q_i$ into $Y_{i,1}$ gives:

$$
Y_{i,1} = \delta_0 + \delta_1 (X_i - 10) + u_i
$$

Expanding the brackets and rearranging:

$$
Y_{i,1} = (\delta_0 - 10\delta_1) + \delta_1 X_i + u_i
$$

Comparing with $Y_{i,1} = \beta_0 + \beta_1 X_i + u_i$ implies:

$$
\delta_1 = \beta_1
$$

Therefore, the OLS estimators satisfy $\hat{\delta}_1 = \hat{\beta}_1$. Thus, while the slope remains unchanged ($\hat{\delta}_1 = \hat{\beta}_1$), the intercept shifts such that $$\hat{\delta}_0 = \hat{\beta}_0 + 10\hat{\beta}_1$$.



**What happens to the OLS coefficients, when a constant is added or subtracted?**

As seen above, the slope coefficient does not change ($\hat{\delta}_1 = \hat{\beta}_1$).

To see how the intercept changes, substitute $Q_i = X_i + c$ into the model:

$$
Y_{i,1} = \delta_0 + \delta_1 (X_i + c) + u_i
$$

Rearranging:

$$
Y_{i,1} = (\delta_0 + c \delta_1) + \delta_1 X_i + u_i
$$

Comparing with $Y_i = \beta_0 + \beta_1 X_i + u_i$ yields:

$$
\beta_0 = \delta_0 + c \delta_1
$$

Solving for the intercept:

$$
\delta_0 = \beta_0 - c \beta_1
$$

Thus, adding or subtracting a constant from a regressor does not affect the slope, and the relationship between $\delta_0$ and $\beta_0$ remains given by $\delta_0 = \beta_0 - c \beta_1$, where $c$ is an arbitrary constant.

## Exercise 2

#### (i) Importing, exploring and cleaning the dataset

```{r, results='hide'}
df <- read.csv("distances_birthtown_up.csv")
skim_without_charts(df)

``` 
In `p0` (minimum value), we see values such as -999999 in`income`, `income_parent1` and `income_parent2` which need to be replaced by `NA`. Also in the variable `travel_time` there are some negative values, which can't be real we also replace them with `NA`.
```{r, results='hide'}
df$income[df$income == -9999999] <- NA
df$income_parent1[df$income_parent1 == -9999999] <- NA
df$income_parent2[df$income_parent2 == -9999999] <- NA

df$travel_time[df$travel_time<0] <- NA
#We create the income_parents variable
df$income_parents <- df$income_parent1 + df$income_parent2
```

<br>

#### (ii) Writing the Regression Equation
We write a regression equation to jointly investigate the relationship of medicine (as the outcome variable) with travel time and income parents (as the independent
variables):

$$medicine_i= \beta_0 + \beta_1travel\_time_i + \beta_2 income\_parents_i + u_i$$

**Interpretation of the coefficient parameters**

Since medicine is a binary variable, this can be interpreted as a linear probability model.

- $\hat{\beta}_0$ (Intercept): The predicted probability of working in the medical sector for an individual with zero travel time and zero parental income; however, this scenario is not economically meaningful, as zero values for these variables are rather unrealistic.
- $\hat{\beta}_1$ (Travel time): A one-hour increase in travel time to Country B from town of birth is associated with a change of $\hat{\beta}_1$ percentage points in the probability of working in medicine, holding parental income constant.
- $\hat{\beta}_2$ (Income parents): A one-unit ($1,000) increase in combined parental income is associated with a $\hat{\beta}_2$ change in the probability of working in medicine, holding travel time constant.
<br><br>
We assume that MLR.1-MLR.5 hold.

<br>

#### (iii) Comparison of the two regressors
From the previous sub exercise, this model accounts for both travel time and parental income:

$$
medicine_i = \beta_0 + \beta_1 \, travel\_time_i + \beta_2 \, income\_parents_i + u_i
$$

From the current sub exercise, this model focuses solely on the relationship with parental income:

$$
medicine_i = \delta_0 + \delta_1 \, income\_parents_i + v_i
$$
The estimation of the two models yield:
```{r}
# Converting medicine to binary from text string
df$medicine_bin <- ifelse(df$medicine == "Yes", 1, 0)

model1 <- lm(medicine_bin ~ travel_time + income_parents, data = df)
model2 <- lm(medicine_bin ~ income_parents, data = df)

summary(model1)
summary(model2)
```

**Comparison**

Let us now compare the point estimates from our specification in part (ii) to those from
(iii):

|                              | Model (ex. ii) | Model (ex. iii) |
|:-----------------------------|:----------------------:|:--------------------------:|
| Coefficient (income_parents) | $\hat{\beta}_2 \approx 0.0002$ | $\hat{\delta}_1 \approx 0.0008$ |
| p-value                      | 0.141                  | < 0.001                    |
| Statistically significant ($\alpha=5\%$) | No                     | Yes                        |
| Economic relevance           | Small                  | Substantially larger (~4x) |


This pattern is consistent with omitted variable bias. In the full model (ex. ii), `travel_time` is included and captures its negative effect on the probability of working in medicine. In contrast, in model (ex. iii), `travel_time` is omitted and only `income_parents` is included. 

Assuming that `travel_time` is negatively correlated with `income_parents`, i.e., $\text{Cov}(travel\_time, income\_parents) < 0,$ omitting `travel_time` in model (ex. iii) leads to an upward bias in the estimated coefficient on `income_parents`. This explains why the coefficient is larger in the simple model (ex. iii).

<br>

#### (iv) Is `region` a better variable than `travel_time`?

Using `region` instead of `travel_time` would likely be less informative for our research question. `travel_time` is a continuous variable that directly measures proximity to Country B's high-income medical industry, capturing meaningful variation across individuals. 

By contrast, `region` is a categorical variable (codes 1–15) that groups individuals into broad geographic areas, thereby discarding within-region variation in actual travel times—two individuals in the same region may face very different travel times to Country B. Furthermore, `travel_time` has a direct economic interpretation (each additional hour is associated with an estimated 4.9 percentage point decrease in the probability of working in medicine), whereas region codes have no inherent ordering or magnitude. 

While `region` may capture broader regional characteristics, such as local infrastructure or access to medical facilities, which could also influence career choices, these factors are not directly observed and `region` serves only as a coarse proxy for them. Therefore, `travel_time` remains a more precise and interpretable measure of geographic proximity to the medical labour market.

<br>

#### (v) Test whether `income_parents` is correlated with `travel_time`


To test whether `travel_time` and `income_parents` are correlated, we estimate the following regression:

$$
income\_parents_i = \gamma_0 + \gamma_1 travel\_time_i + e_i
$$

The null and alternative hypotheses are:

- $H_0$: $\gamma_1 = 0$ (no correlation)  
- $H_1$: $\gamma_1 \neq 0$ (correlation exists)

The t-statistic is:
$$
t = \frac{\hat{\gamma}_1}{SE(\hat{\gamma}_1)}
$$
where \(SE(\hat{\gamma}_1)\) is the standard error of the estimated coefficient \(\hat{\gamma}_1\), measuring the sampling variability of the estimate.

We conduct the test at the 5% significance level (\(\alpha = 0.05\)).

```{r}
# Estimate model
model_corr <- lm(income_parents ~ travel_time, data = df)

# Extract coefficient and standard error
beta_hat <- coef(model_corr)["travel_time"]
se_beta  <- summary(model_corr)$coefficients["travel_time", "Std. Error"]

# Compute t-statistic manually
t_stat <- beta_hat / se_beta

# Degrees of freedom
n <- nrow(df)
k <- length(coef(model_corr)) - 1
df_t <- n - k - 1

# Critical value (two-sided, 5%)
t_crit <- qt(0.975, df_t)

# Output
t_stat
t_crit

```

```{r}
# Decision rule
if(abs(t_stat) > t_crit){
  "Reject H0"
} else {
  "Do not reject H0"
}
```

The computed t-statistic is compared to the critical value from the t-distribution table.

Since \(|t| = 58.52 > t_{crit} \approx 1.96\) (for a two-sided test at the 5\% significance level and large degrees of freedom), we reject \(H_0\) at the 5\% significance level. This implies that `travel_time` and `income_parents` are statistically significantly negatively correlated.


<br>

#### (vi) Test whether `travel_time` and `income_parents` are jointly significant

We consider the regression model:

$$
medicine_i = \beta_0 + \beta_1 \, travel\_time_i + \beta_2 \, income\_parents_i + u_i
$$
The null and alternative hypotheses are:

- $H_0$: $\beta_{1} = 0$ and $\beta_{2} = 0$  
  (no joint effect)

- $H_1$: $\beta_{1} \neq 0 \ \text{or} \ \beta_{2} \neq 0$  
  (at least one coefficient is different from zero, i.e. joint significance)

We use an F-test comparing the unrestricted model (including both variables) to the restricted model (excluding them).

The test statistic is:
$$
F = \frac{(SSR_r - SSR_{ur})/q}{SSR_{ur}/(n - k - 1)}
$$
where \(q\) is the number of restrictions. In this case, \(q = 2\), since both coefficients are tested jointly.

We conduct the test at the 5% significance level (\(\alpha = 0.05\)). The critical value \(F_{crit}\) is obtained from the F-distribution table provided in the assignment, using \(q = 2\) numerator and \(n - k - 1\) denominator degrees of freedom.

To implement the test, we estimate both models and compute the F-statistic manually in R using the residual sum of squares (SSR). 


```{r}
# Unrestricted model: includes both variables of interest
model_ur <- lm(medicine_bin ~ travel_time + income_parents, data = df)

# Restricted model: excludes both variables (only intercept)
model_r  <- lm(medicine_bin ~ 1, data = df)

# Compute residual sum of squares (SSR) for both models
SSR_ur <- sum(resid(model_ur)^2)
SSR_r  <- sum(resid(model_r)^2)

# Number of restrictions
q <- 2

# Sample size
n <- nrow(df)

# Number of regressors in unrestricted model (excluding intercept)
k <- length(coef(model_ur)) - 1

# Compute F-statistic manually
F_stat <- ((SSR_r - SSR_ur)/q) / (SSR_ur/(n - k - 1))

# Critical value at 5% significance level
F_crit <- qf(0.95, q, n - k - 1)

# Output results
F_stat
F_crit

```

```{r}
# Decision rule
if(F_stat > F_crit){
  "Reject H0"
} else {
  "Do not reject H0"
}
```

<br>

The computed F-statistic is 306.45, which is substantially larger than the critical value of 2.996 at the 5% significance level.

Therefore, we reject the null hypothesis \(H_0\) at the 5% significance level. This implies that at least one of the coefficients on `travel_time` or `income_parents` is different from zero, i.e. the variables `travel_time` and `income_parents` are jointly statistically significant.



<br>
