---
title: "Assignment_2_Statistics_II"
author:
- "Maik David Klaiber (22-482-426)"
- "Jonas Weber (24-621-484)"
- "Yann Rochaix (24-622-276)"
date: "`r format(Sys.Date(), '%d.%m.%Y')`"
output:
  html_document: default
  pdf_document: default
---
```{r setup, include =FALSE}
library(tidyverse)
library(summarytools)
library(skimr)
```
## Exercise 1 - Data Preparation
We imported the datasets using `read_delim()` and `read_csv()` from the `readr` package to handle different separators. We then used `skim_without_charts` from the `skimr` package to automatically check the data without the graphics. The difference in row counts gives us a first hint, that there might be a mismatch in the number of cantons. This summary also shows us, that there are no missing values. 
```{r, message=FALSE, results='hide'}
cantons<- read_delim("ps2_cantons_cleaned.csv", delim = ";")
surnames<- read_csv("ps2_cantons_surnames.csv")
skim_without_charts(cantons)
skim_without_charts(surnames)
```
We used `mutate()` to consolidate the individual proportions into a single surname variable and then applied `select()` to remove the original columns, streamlining the tibble for a cleaner merge. Then, to show the rows that have no  matching canton name in the canton dataset, we use `antijoin` to reveal them. 
```{r, message=FALSE}
surnames <- surnames %>%
  mutate(surname = share_mueller + share_maier) %>%
  select(-share_mueller, -share_maier)

surnames %>% 
  anti_join(cantons, by = "canton")
```
In this step, we harmonized the naming convention by recoding "Sankt Gallen" to "St. Gallen" in order to match in both datasets and filtered out the "Sri Lanka" entry to ensure the surnames dataset exclusively contains valid Swiss cantons for the subsequent join. After performing the merge, we used `select()` to retain only the essential variables for analysis: GDP, population, immigration, and the newly calculated surname shares. We then verified the integrity of the join by confirming there are zero missing values and that all 26 distinct Swiss cantons are correctly represented in the final data frame.
```{r}
surnames <- surnames %>%
  mutate(canton = recode(canton, "Sankt Gallen" = "St. Gallen")) %>%
  filter(canton != "Sri Lanka")

df <- cantons %>% #Merge the surname dataset into the canton dataset
  left_join(surnames, by = "canton") %>%
  select(gdp, population, immigration, surname)

sum(is.na(df$surname))          
n_distinct(cantons$canton)
```
## Exercise 2 - Estimation of an univariate model
The specific regression equation for this problem is:
$$gdp_i = \beta_0 + \beta_1 \cdot surname_i + u_i$$
```{r}
model <- lm(gdp ~ surname, data = df)
summary(model)

```
### Results
**Intercept $\hat{\beta}_0 = 10,077.28$:** A canton with a `surname` share of zero is predicted to have a GDP of CHF 10,077 million. This is not meaningful in practice since all cantons have some share of these surnames.

**Slope $\hat{\beta}_1 = 449,550$:** A 1 percentage point increase (0,01 unit) in the combined Mueller/Maier surname share is associated with a predicted GDP increase of CHF 4'495,50 million. This reflects that larger cantons naturally have both more people with common surnames and higher total GDP, so the relationship mainly captures canton size rather than a causal effect of surnames.


We use the average `surname` share across all 26 cantons to predict GDP at a typical canton-level surname share.
```{r}
avg_surname <- mean(df$surname)
predicted_gdp <- predict(model, newdata = data.frame(surname = avg_surname))
print(as.numeric(predicted_gdp))
```

For each additional percentage point increase in the Mueller/Maier surname share beyond this average, GDP is predicted to increase by CHF 4,495.50 million.


## Exercise 3 - Comparison of two univariate models

We estimate the regression of $gdp$ on $\log(\text{immigration})$ and compute the corresponding $R^2$.
```{r}

# Estimate regression of GDP on the log of immigration
model_log <- lm(gdp ~ log(immigration), data = df)
summary(model_log)

# Define R² from the population model estimated in Exercise 2
r_squared_surname <- summary(model)$r.squared

# Create a small table comparing the R² values of both models
r2_table <- tibble(
  model = c("surname (exercise 2)", "log(immigration) model"),
  R_squared = c(r_squared_surname, summary(model_log)$r.squared)
)

r2_table

```

<b>What does $R^2$ measure here intuitively?</b><br>
$R^2$ measures the proportion of the variation in $y$ that is explained by the regression model. A higher $R^2$ indicates greater explanatory power of the model.

<b>What amount of variation in `gdp` is explained?</b><br>
The surname model explains about 64.3\% of the variation in $gdp$, while the log(immigration) model explains about 60.5\%. Thus, the surname model has higher explanatory power.

<b>Unbiasedness of the Estimators for Surname and Immigration</b><br>
The OLS estimators are unbiased if the assumption $E(u \mid X) = 0$ holds (SLR.4), meaning that the error term has an average value of zero for any given value of $X$. This requires that the explanatory variable is not systematically related to omitted factors that affect $gdp$.

In the surname model, this assumption is likely violated because surname shares may be correlated with underlying regional or demographic characteristics that also influence economic performance. If such relevant factors are omitted from the regression, the estimator for the surname coefficient may be biased due to omitted variable bias.

A similar issue may arise in the immigration model. For example, `education levels` could be an omitted variable. Higher education levels can increase productivity and therefore $gdp$. At the same time, economically attractive cantons with better labor market opportunities may attract more immigrants. If education is correlated with immigration and affects $gdp$, the immigration coefficient may capture part of this effect, leading to biased OLS estimates.

<b>Relationship between unbiasedness of a coefficient estimator and $R^2$</b><br>
Unbiasedness of an estimator depends on the assumption $E(u \mid X) = 0$. In contrast, $R^2$ measures how much of the variation in the dependent variable is explained by the model. Therefore, unbiasedness and predictive power are not directly related and a model can have a high $R^2$ while the coefficient estimates are still biased.


## Exercise 4 - Univariate model without a constant

We state the formula without the constant $\beta_0$ as $$
gdp_i = \beta_1 \, surname_i + u_i
$$

```{r}

# Estimate the regression of gdp on surname without an intercept (-1)
lm(gdp ~ surname - 1, data = df)

```

<b>Why did the coefficient on surname change compared to subexercise 2?</b><br>
The coefficient changes because the regression is estimated without an intercept, forcing the regression line to pass through the origin. In this case, the OLS estimator simplifies from $\hat{\beta}_1 = \frac{\sum (x_i-\bar{x})(y_i-\bar{y})}{\sum (x_i-\bar{x})^2}$ to $\hat{\beta}_1 = \frac{\sum x_i y_i}{\sum x_i^2}$, which changes the estimated slope (from 449,550 to 532,247).


## Exercise 5 - Code your own OLS estimator

We estimate the parameters of the univariate regression model

$$
y_i = \beta_0 + \beta_1 x_i + u_i .
$$

Using the OLS formulas, the slope coefficient is given by

$$
\hat{\beta}_1 =
\frac{\sum_{i=1}^{N}(x_i-\bar{x})(y_i-\bar{y})}
{\sum_{i=1}^{N}(x_i-\bar{x})^2}.
$$

The intercept is then calculated as

$$
\hat{\beta}_0 = \bar{y} - \hat{\beta}_1 \bar{x}.
$$

In our application, we regress `gdp` (as Y) on `surname` (as X), including an intercept. 
Thus, the vectors `gdp` and `surname` are used as inputs to compute the OLS estimators using the formulas above.

```{r}

estimate_ols <- function(y, x) {
  
  # Compute the OLS slope coefficient beta_1 using the analytical formula above
  beta_1 <- sum((x - mean(x)) * (y - mean(y))) / sum((x - mean(x))^2)
  
  # Compute the intercept beta_0
  beta_0 <- mean(y) - beta_1 * mean(x)
  
  # Return the estimated coefficients as a named vector
  return(c(beta_0 = beta_0, beta_1 = beta_1))
}

estimate_ols(df$gdp, df$surname)

```

<b>Verification against the lm() function</b><br>
The manually computed coefficients are virtually identical to those obtained with `lm()` in Exercise 2. This confirms that our implementation correctly reproduces the OLS estimators.


<b>Modification of function</b><br>
We extend our OLS function by adding the boolean argument `add_constant`, which allows us to estimate the regression either with or without an intercept.

```{r}

    #In this case "false" -> in order to verify beta_1 with the exercise 3 above
estimate_ols_modified <- function(y, x, add_constant = FALSE) { 
  
  if (add_constant == TRUE) {
    
    # OLS with intercept
    beta_1 <- sum((x - mean(x)) * (y - mean(y))) / sum((x - mean(x))^2)
    beta_0 <- mean(y) - beta_1 * mean(x)
    
    return(c(beta_0 = beta_0, beta_1 = beta_1))
    
  } else {
    
    # OLS without intercept
    beta_1 <- sum(x * y) / sum(x^2)
    
    return(beta_1)
  }
}

estimate_ols_modified(df$gdp, df$surname)

```

<b>Verification with result in subexercise 4</b><br>
The estimated coefficient obtained with `add_constant = FALSE` is identical to the result from `lm(gdp ~ surname - 1, data = df)` in subexercise 4 (532247), confirming that the implementation correctly reproduces the OLS estimator for the model without an intercept.

<br>
<br>








